
###--------------------------------###
#Module 04: Data Science in R ####
#MB5370
#Nick Murray 2/5/2023
---------------------------------
```{r}
setwd("C:/Users/John/OneDrive/Desktop/Semester 3/MB5370 (Techniques in Marine Science 1/Module 4-Data Science")
```

#Install the tidyverse
# install.packages("tidyverse")
#(librarytidyverse)

# ---------------------
##2.6 Create your first ggplot

#plot the data (data=mpg)
#then the size of the engine (displ) on the x axis
#then the fuel efficiency on the y axis

```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy))
```
###2.7 Understand the 'Grammar of grpahics
##2.7.1 Graphing Template
```{r}
ggplot(data = <DATA>) + <GEOM_FUNCTION> (mapping = aes(<MAPPINGS>))
```
##2.7.2 Aesthetic mappings
#when creating a plot you essentially need two attributes of a plot: a geom and aesthetics
#add the colour to the graph
```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy, colour = class))

```
#can change point size by class
#size of the points now represent the class of car
#unwise and looks ugly

```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy, size = class))
```
#change point transparancy by class (alpha)
#alpha dictates transparancy 
```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy, alpha = class))
```
#Change point shape by class
```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy, shape = class))
```
#can also set these properties manually, by offering a number or a colour
#make all the points blue
#changes appearance of the dot it doesnt show anything about the nature of the variable
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy), color = "blue")
```
#range of manual aesthetics you can set to manually alter the appearance
# the name of a color as a character string
#the size of a point in mm
#the shape of a point as a number
```{r}
ggplot(data=mpg)+ geom_point(mapping=aes(x=displ,y=hwy), colour = 5)
```
### 2.8 TroubleShooting

### 2.9 facet and Panel Plots

#to break up your plot into sub plots use the facets
#to facet your plot use facet_wrap()
#only use facet_wrap() for discrete variables
```{r}
ggplot(data=mpg)+
  geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)
```
#if you want to do this with more than one variable use facet_grid() need to variables using ~ to split them up
#use a . if you do not want to facet in the rowa or column dimension
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy)) + facet_grid(drv ~ cyl)
```
#use a . if you do not want to facet in the row or column dimension
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy)) + facet_grid(. ~ cyl)
```
###2.10 Fitting simple lines
#to display data as points, use geom_point()
# but ggplot can use a variety of geom objects to represent data
#geom is an object that your plot uses to represent the data
```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy))
```
#can comment and test different geom_points out and then comment them out
```{r}
ggplot(data = mpg) +
geom_point (mapping = aes(x= displ, y = hwy)) + #points horribl
geom_smooth(mapping = aes(x = displ, y = hwy))
```

#can have different types of lines for geom_smooth
```{r}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```
###2.11 Transformations and Stats
##2.11.1 plotting statistics
```{r}
ggplot(data = diamonds) + geom_bar(mapping = aes (x = cut))
```
# can generally use geoms and stats interchangeably, can substitute stat_count() instead of geom_bar()
```{r}
ggplot(data = diamonds) + stat_count(mapping = aes(x = cut))
```
##2.11.2 overriding defaults in ggplot2
#may want to overrride a default stat
```{r}
demo <- tribble (
  ~cut,   ~freq,
  "Fair",  1610,
  "Good",  4906,
  "Very Good", 12082,
  "Premium",   13791,
  "Ideal",  21551
)
demo
ggplot(data = demo) +
  geom_bar(mapping = aes (x=cut, y= freq), stat = "identity")
```
##2.11.3 plotting statistical details
#might want to show a little more about transformations in your plot
#can do this using stat_summary()
```{r}
ggplot(data = diamonds) +
  stat_summary(
    mapping = aes (x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```
# in our proportion bar chart, we need to set group = 1 why? what is the problem with these two graphs
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x=cut, y= after_stat(prop)))
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y= after_stat(prop)))
```
###2.12 Positional Adjustments
# to boost the way you can convey information with plots, you can use colour or fill to change aspects of bars
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
#now try using these aesthetics to colour by another variable like clarity
#the ability to make position adjustments is vital can customize your plots in three ways
# identity (raw data), fill (changes heights) and dodge (which forces ggplot2 to not put things on top of eachother)
# if you use position = "identity",you will be able to place each object exactly where it falls in the context of the graph
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
```{r}
#to alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```
```{r}
#to color the bar outlines with no filler color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
#Position = "fill" works like stacking, but each of the set of stacked bars the same height
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position ="fill")
```
#Position = "dodge places overlapping objects directly beside one another
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```
# use position jitterto avoid overplotting when points overlap. useful for scatterplots but not barplots
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```
#above we learned how to plot almost any function,
#this is the recipe for almost any type of plot you want
```{r}
#ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(
#     mapping = aes(<MAPPINGS>),
#     stat = <STAT>, 
#     position = <POSITION>
#  ) +
#  <FACET_FUNCTION>
```
###2.14 (Homework plot deconstruction)
#----------------------------------------#

####Workshop 2 Using ggplot2 for communication####-----------

###3.1 Labels

#add labels with the labs() function
#start with title
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```
#if you need to add more text You can use functions:
# subtitle- adds smaller detail text below the title
#caption- adds text at the bottom right of the plot, often used to describe the source of the data
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```
#can also use labs() to replace axis labels abd legend titles
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) + 
  geom_smooth(se = FALSE) + 
  labs(
    x = "Engine Displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car Type"
  )
```
###3.2 Annotations --------------------

#can use geom_text() to add textual labels to your plots
#allows us to pull data values out from the data itself and include it in the plot
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class) +
  labs(
    x ="Engine Displacement(L)" ,
    y = "Highway Fuel (mpg)"
  )
```
###3.3 Scales----------------
#ggplot automatically adds scales behind the scenes
#if you want to customize them add number in the appropriate scale arguments
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

###3.4 Axis ticks
#breaks control the postion of the ticks and can use labels as an argument to change text label
#seq(15, 40, by = 5)) makes the graph start at 15 and end at 40 in increments of 5
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```
#similarly you can use NULL to suppress labels altogether 
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```
### 3.5 Legends and Colour Schemes--------------
#if you want to chang ethe overrall position of the legend you need to use theme() 
#can also use legen.position = "none" to get rid of legend
```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right")
```
###3.6 Replacing scale---------------#
#two types of scales you would want to switch out: continuous position scales and colour scale
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10(1) + 
  scale_y_log10(1)
```
# another scale that is customized is colour scale
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
ggsave("my-plot.pdf")
```
#when you have predefined colours you want to set yourself, can set them using
# scale_colour_manual()
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```
#one of the best colours out there is viridis
```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```
### 3.7 Themes-----------#
#now you can customize the entire theme of your plot
#themes are all of the non-data elements of your plot
#ggplot2 has 8 themes by default, many more in ggthemes
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```
#example of nicks theme so his plots all look the same
```{r}
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```

###3.8 Saving and exporting your plots
# best way to make a single file will be to export it using the ggsave() function
#will save your most recent plot to your working directory
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
```

####Workshop 3 Reproducible Work---------#

```{r}
install.packages("usethis")
library(usethis)
use_git_config(user.name = "JohnAbnet11", user.email = "john.abnet1@gmail.com")
```


####Workshop 4 Data Wrangling

###5.3 What is a tibble

#Tibbles are a type of data frame 
#they help future proof code in an aging coding langauge
#tibble is part of tidyverse package
#to convert regular data frames into tibbles you can use
```{r}
as_tibble()
```

#also tribbles which stand for transposed tibble (tribble())
#tribbles have one purpose: to help you do data entry directly in your script
```{r}
tribble(
  ~x , ~y, ~z,
  #--
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

#lets look at a difference between a tibble and a data.frame
#tibbles only print first 10 rows and all columns that can fit on the screen, easier with large data sets

#lets have a look at examining data using tibble
```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
#> # A tibble: 1,000 × 5
#>   a                   b              c     d e    
#>   <dttm>              <date>     <int> <dbl> <chr>
#> 1 2022-11-18 23:15:18 2022-11-25     1 0.368 n    
#> 2 2022-11-19 17:20:28 2022-11-30     2 0.612 l    
#> 3 2022-11-19 11:44:07 2022-12-10     3 0.415 p    
#> 4 2022-11-19 01:05:24 2022-12-09     4 0.212 m    
#> 5 2022-11-18 21:29:41 2022-12-06     5 0.733 i    
#> 6 2022-11-19 08:30:38 2022-12-02     6 0.460 n    
#> # ... with 994 more rows
```
#may need to display more than top few rows and with tibbles you can use 
#print() to designate the number of rows (n) and display width (width = Inf #displays every column)

#now lets learn how to pull out a single variable from our dataframe.
#can use a dollar sign ($) to extract a full column of data (by name) or
#the simple double bracket to pull out an exact row of data (by row position). 
#pull out an exact cell value using [[,]]
```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)
#extract by name
df$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df[["x"]]
# [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df[[1]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
#> 
# Extract by exact position
df[[2,2]]

```
#why would you want to do this ?
# could be to add a label to a plot

#later on we will start to use pipes (%>%) which is a new way to do things to variables in R
#for now just know that to use these base R function we will need to use a . as a placeholder when using pipes
```{r}
df %>% .$x
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
df %>% .[["x"]]
#> [1] 0.73296674 0.23436542 0.66035540 0.03285612 0.46049161
```

#if tibbles wont do partial matching, if variable you call out doesnt match exactly whats in the data frame tibbles will generate a warning

### 5.4 How can I import data

#read_csv() reads commas delimited files, read_csv2() reads semicolon sperated files

#the first and most important argument to read_csv() is the file

#the first and most important argument to read_csv() is the file path
#read_csv() uses the first line of data for column names but there are some cases where you wont want the very first line in order to skip the first line you can do skip = n where n is the number of lines you want to skip

#can also change the defaults if your data does not contian column names; use col_names = FALSE to tell read_csv() not to treat the first row as headings

#VERY IMPORTANT can also set the no data values

### 5.5 Tidying data using Tidyr

#table1 is "tidy" once we look at the differences of these we will figure out what makes something tidy
table1
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

# Spread across two tibbles
table4a  # cases
#> # A tibble: 3 × 3
#>   country     `1999` `2000`
#> * <chr>        <int>  <int>
#> 1 Afghanistan    745   2666
#> 2 Brazil       37737  80488
#> 3 China       212258 213766
table4b  # population
#> # A tibble: 3 × 3
#>   country         `1999`     `2000`
#> * <chr>            <int>      <int>
#> 1 Afghanistan   19987071   20595360
#> 2 Brazil       172006362  174504898
#> 3 China       1272915272 1280428583
  
#how we make our dataset a tidy dataset is by following three interrelated rules
#1. Each variable must hav eits own column
#2 eac hobservation must have its own row
#3 Each value must have its own cell

#these three rules leads to an even simpler set of practical instructions:
#1 put each dataset in a tibble
#2 put each varaible in a column

##5.5.2 Spreading and gathering data tables

#first step in tidying the data is to understand what each variable and observation actually means.

#once you understand the data you are looking at, the second step is to resolve one of the two common problems with untidy data:
#1. One variable is spread across multiple columns
#2 one observation is scattered across multiple rows

#to fix these we will explore the use of two functions in tidyr:
# pivot_longer()
#pivot_wider ()

#Using pivot_longer, according to R4DS
#common problem is a dataset where some of the column names are not names of variables
#but values of avariable
#take table4a: the column names 1999 and 2000 represent values of the year variable, the values in the 1999 and 2000 columns represent values of the cases variable
```{r}
#> # A tibble: 3 × 3
#>   country     `1999` `2000`
#> * <chr>        <int>  <int>
#> 1 Afghanistan    745   2666
#> 2 Brazil       37737  80488
#> 3 China       212258 213766
```
#to tidy a dataset we need to pivot the offending columns into a new pair of variable
#we need 3 parameters:
#---- the set of columns whose names are values, not varaibles. in this example, those are the columns 1999 and 2000
#---- the name of the variable to move the column names to. here it is year
#---- the name of the variable to move the column values to. Here it is cases
#these parameters call for the pivot_longer():
```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
#> # A tibble: 6 × 3
#>   country     year   cases
#>   <chr>       <chr>  <int>
#> 1 Afghanistan 1999     745
#> 2 Afghanistan 2000    2666
#> 3 Brazil      1999   37737
#> 4 Brazil      2000   80488
#> 5 China       1999  212258
#> 6 China       2000  213766
```

#columns to pivot are specified with dplyr:: select() style notation.
#only two columns so we list them individually 
#1999 and 2000 are non-syntactic( dont start with a letter) so we have to surround them in backticks year and cases do nto exist in table4a so we put their names in quotes

#pivot_longer() makes datasets "longer" by increasing the number of rows and decreasing the number of columns

------

#using pivot_wider()
# to handle an observation scattered across multiple rows. pivot_wider() is the opposite of pivot_longer() in table2 we see an observation is a country in a year
#with the observation spread across two rows:
```{r}
table2
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
```

#to make this tidy we only need two parameters:
#1 the column to take variables from: type
#2 the column to take values from: count

#now we can use pivot_wider() to make the table shorter and wider by creating ne separate columns for cases and population and populating them with their associated values

### 5.5.3 Separating and uniting data tables

# have tidied table2 and table 4 nut now we will explore table3 which has a different issue that makes it untidy

#in table3 we see one column (rate) that contains two variables (cases and population)
#can use the separate() function which separates one column into multiple columns
```{r}
table3
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583
```


```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
#> # A tibble: 6 × 4
#>   country      year cases  population
#>   <chr>       <int> <chr>  <chr>     
#> 1 Afghanistan  1999 745    19987071  
#> 2 Afghanistan  2000 2666   20595360  
#> 3 Brazil       1999 37737  172006362 
#> 4 Brazil       2000 80488  174504898 
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```
 
 # usually splits table by non numerical character but can specify by using 
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```
 # both cases and population are listed as character types
 #we want them to be numbers so we use convert = TRUE
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```
 
### 5.5.4 Handling Missing values

#### Workshop 5------------------------------#

### 6.4 Installing the Spatial R packages

```{r}
#install necessary packages
install.packages("sf")
install.packages("terra")
install.packages("leaflet")
install.packages("tmap")
#load into R library
library(tidyverse)
library(sf)
library (terra)
library(leaflet)
library(tmap)
library(mgcv)
```
### 6.6 loading the spatial dataset

```{r}
library(readr)
setwd("C:/github/MB5370")
dat <- read_csv("Data/copepods_raw.csv")
dat
```
#First we must understand what this data is telling us

### 6.7.1 Check Coordinates
```{r}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```
#this looks good, but this is not a map, it doesnt have critical things a real map needs, and real distances between these points are probably not correct

```{r}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```

### Getting going with maps

#we will repeat the map of richness, but now using some of Rs specilist packages
#turn our point data into a spatially referenced data frame using sf package
```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
#st_as_sf converts different data types to simple features
#dat is our orginal data
#coords gives the name of the columns that relate to the spatial coordinates of X coordinate
#crs stands for coordinate reference system which we will discuss next
```

### 6.9 Coordinate reference systems

```{r}
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name # pull out just the name of the crs
[1] "WGS 84"
```
```{r}
crs4326$wkt # crs in well-known text format
```

### 6.10 Feature Collection(Points)

```{r}
sdat
```

### 6.11 Cartography

```{r}
plot(sdat["richness_raw"])
```
#we have only plotted the richness column

```{r}
plot(sdat)
```

### 6.12 Thematic maps for communication
#will use tmap to build and add on layers
#here we only have one layer from sdat
#we declare the layer with tm_shape()
```{r}
tm1 = tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
```

#use tmap_save to save the map to your working directory
```{r}
tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)
```

###6.13 Mapping spatial polygons as layers

##6.13.1 Loading shapefiles

```{r}
setwd("C:/github/MB5370")
aus <- st_read("Data/Aussie.shp")
```
## 6.13.2

